---
title: "Survival Analysis of Diabetes Incidence"
author: "Lauren Bastian"
date: "`r format(Sys.time(), '%m/%d/%y')`"
output: 
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup images, include = FALSE}
knitr::opts_chunk$set(dev = "png",
                      dpi = 300,
                      echo = FALSE,
                      cache = TRUE)
```

# Study of Women's Health Across the Nation

The data come from the Study of Women's Health Across the Nation (SWAN), a longitudinal population-based study of 3302 middle-aged women with data collected between 1998 and 2008, with data made public for research purposes through [ICPSR](https://www.icpsr.umich.edu/web/ICPSR/series/00253). The analytical sample contains 2686 records of middle-aged women who participated in the study and had complete observations for race, baseline BMI, baseline household income, and health insurance. Race, household income, and insurance status were assessed during a baseline visit and are assumed to be indicative of race, household income, and insurance status throughout the study. Diabetes status was assessed during baseline as whether or not the individual was currently treating diabetes using insulin, and explicitly assessed during ten follow-up visits ($t$ is the years since baseline visit). 

```{r libraries and dataset, include = FALSE, fig.show = "hide", results = "hide"}
##SET WORKING DIRECTORY TO SOURCE FILE LOCATION

##LIBRARIES
#install.packages("survival")
#install.pacakges("ggplot2)
#install.packages("survminer")
#install.packages("rms")
#install.packages("gmodels")
#install.packages("lsr")
library(survival)
library(ggplot2)
library(survminer)
library(rms)
library(gmodels)
library(lsr)

##LOAD DATA
source("build_survival_dataset_updated.R")

surv.df$race = as.factor(surv.df$race)
levels(surv.df$race) = c("Black/African American",
                         "Chinese/Chinese American",
                         "Japanese/Japanese American",
                         "Non-Hispanic White/Caucausian",
                         "Hispanic")

surv.df$income = as.factor(surv.df$income)
surv.df$income = factor(surv.df$income, 
                        levels = c("<20","20-49","50-99","100+"))

surv.df$age_group = as.factor(surv.df$age_group)

surv.df$bmi_group = as.factor(surv.df$bmi_group)

surv.df$insured = as.factor(surv.df$insured)

surv.df$event = as.numeric(surv.df$event)

##Create surv object in surv.df
surv.df$surv_obj = with(surv.df, Surv(t2event, event))

attach(surv.df)

```


# Research Objectives

How does insurance status affect the onset of diabetes among middle-aged American women over 10 years? Does this affect remain after controlling for race and household income, BMI, and age group?

How do race and income affect onset of diabetes among middle-aged American women? Does the affect of race on diabetes onset remain after controlling for household income, BMI, and age group?

How does yearly household income, BMI, race, and insurance affect the onset of diabetes among the middle-aged American women? Are there any interactions between these factors?

# Descriptive Statistics

Histogram of time to event for uncensored individuals. Time to event is left skewed with a peak at 10 years.  

```{r Hist_t2event_IncidentDiabetes, echo = FALSE}
windowsFonts(Times=windowsFont("TT Times New Roman"))

event.hist = ggplot(data = surv.df[surv.df$censored == 0,], aes(x = t2event)) +
  geom_histogram(binwidth = 1, fill = '#00BFC4', color = 'white') +
  theme_bw() +
  labs(title = "Time to Event for Individuals w/ Incident Diabetes") +
  xlab("Time to Event") +
  ylab("Counts") +
  geom_vline(xintercept = mean(surv.df[surv.df$censored == 0,]$t2event), size = 1, linetype = "longdash") +
  theme(text = element_text(family = "Times"))

event.hist
```
```{r}
mean(surv.df[surv.df$censored == 0,]$t2event)
sd(surv.df[surv.df$censored == 0,]$t2event)
```

```{r Hist_Age_IncidentDiabetes, echo = FALSE}
windowsFonts(Times=windowsFont("TT Times New Roman"))

event.hist = ggplot(data = surv.df[surv.df$censored == 0,], aes(x = t0_age + t2event)) +
  geom_histogram(binwidth = 1, fill = '#00BFC4', color = 'white') +
  theme_bw() +
  labs(title = "Age of Individuals w/ Incident Diabetes") +
  xlab("Age") +
  ylab("Counts") +
  geom_vline(xintercept = mean(surv.df[surv.df$censored == 0,]$t2event + surv.df[surv.df$censored == 0,]$t0_age), size = 1, linetype = "longdash") +
  theme(text = element_text(family = "Times"))

event.hist
```

```{r}
mean(surv.df[surv.df$censored == 0,]$t0_age + surv.df[surv.df$censored == 0,]$t2event)
sd(surv.df[surv.df$censored == 0,]$t0_age + surv.df[surv.df$censored == 0,]$t2event)
```

```{r, echo = FALSE}
xtabs(~surv.df$race + surv.df$event)
```

```{r, echo = FALSE}
xtabs(~surv.df$bmi_group + surv.df$event)
```

```{r, echo = FALSE}
xtabs(~surv.df$insured + surv.df$event)
```

```{r, echo = FALSE}
xtabs(~surv.df$income + surv.df$event)
```

```{r, echo = FALSE}
xtabs(~surv.df$age_group + surv.df$event)
```

```{r Hist_t2event_allsubjects, echo = FALSE}
all.subjects = surv.df$t2event
event.subjects = surv.df$t2event[surv.df$censored == 0]
all.subjects = data.frame("t2event" = all.subjects, "subset" = rep("All Subjects", 2686))
event.subjects = data.frame("t2event" = event.subjects, "subset" = rep("Event Occured", 274))
hist.data = rbind(all.subjects, event.subjects)


ggplot(data = hist.data, aes(x = t2event, fill = subset)) +
  geom_histogram(position = 'identity', binwidth = 1, alpha = 1, color = 'white') +
  theme_bw() +
  labs(title = "Time to Incident Diabetes for Censored versus All Individuals") +
  xlab("Time to Event") +
  ylab("Counts") +
  theme(text = element_text(family = "Times"), legend.title=element_blank()) +
  scale_fill_manual(values=c("gray85", "#00BFC4"),
                       labels=c("All Individuals", "Uncensored"))

```



The mean age at $t = 0$ is 45.899.
```{r descriptives, echo = FALSE}
##MEAN AGE AT t = 0
mean(surv.df$t0_age)
```
Count distribution of the age group upon study entry. This will be our age strata.

```{r, echo = FALSE}
##AGE GROUP DISTRIBUTION
table(surv.df$age_group)
```

Count distribution of insurance status at baseline. There are 171 uninsured and 2515 insured women in the study.

```{r, echo = FALSE}
##INSURANCE STATUS
#6.65% of participants report no insurance during screening study
table(surv.df$insured)
```

6.37% of women in the study are uninsured at baseline. 

```{r, echo = FALSE}
nrow(surv.df[surv.df$insured == "No",]) / nrow(surv.df) 
```
Race counts of the analytical sample ($n = 2686$)

```{r, echo = FALSE}
##RACE
table(surv.df$race)
```

Distribution of self-reported household income at baseline.

```{r, echo = FALSE}
##HOUSEHOLD INCOME
table(surv.df$income)
```

Histogram of continuous BMI. BMI group cutoff points are indicated by vertical dashed lines.

```{r Hist_BaselineBMI, echo = FALSE}
cutoffs = c(30, 40)

ggplot(surv.df, aes(x = t0_bmi)) +
  geom_histogram(binwidth = 1, fill = '#00BFC4', color = 'white') +
  theme_bw() +
  labs(title = "Baseline BMI") +
  xlab("BMI") +
  ylab("Counts") +
  geom_vline(xintercept = cutoffs, size = 1, linetype = "longdash") +
  theme(text = element_text(family = "Times"))

```

Distribution of BMI groups. There are 186 women in the highest BMI group.

```{r, echo = FALSE}
table(surv.df$bmi_group)
```

Incident diabetes counts over the course of the study. 274 participants out of 2686 developed diabetes during the course of the study, 11.36% of the analytical sample.

```{r, echo = FALSE}
##OVERALL EVENT INCIDENCE DURING STUDY
table(surv.df$event)
274/2412
```


# Exploratory Analysis

Cross tabulation for insurance and income.

```{r}
ftable(xtabs(~insured + income, data = surv.df))

CrossTable(surv.df$insured, surv.df$income,
           prop.chisq = FALSE,
           format = "SPSS",
           digits = 2)
```

Cross tabulation for income and BMI. 

```{r}
ftable(xtabs(~bmi_group + income, data = surv.df))

CrossTable(surv.df$bmi_group, surv.df$income,
           prop.chisq = FALSE,
           format = "SPSS",
           digits = 2)
```

Cross tabulation for race and income. 

```{r, echo = FALSE}
##FACTOR CROSS TABS
ftable(xtabs(~race + income, data = surv.df))

CrossTable(surv.df$race, surv.df$income,
           prop.chisq = FALSE,
           format = "SPSS",
           digits = 2)
```

Cross tabulation for race and insurance status.

```{r, echo = FALSE}
##FACTOR CROSS TABS
ftable(xtabs(~race + insured, data = surv.df))

CrossTable(surv.df$race, surv.df$insured,
           prop.chisq = FALSE,
           format = "SPSS",
           digits = 2)
```

Cross tabulation for BMI and race.

```{r, echo = FALSE}
ftable(xtabs(~race + bmi_group, data = surv.df))

CrossTable(surv.df$race, surv.df$bmi_group,
           prop.chisq = FALSE,
           format = "SPSS",
           digits = 2)
```

Incident diabetes by race.

```{r, echo = FALSE}
#Incident Diabetes BY RACE
xtabs(~race + event, data = surv.df)
```



Incident diabetes by insurance status. 

```{r, echo = FALSE}
##EVENT INCIDENCE BY INSURANCE STATUS
xtabs(~insured + event, data = surv.df)
```



Incident diabetes by household income at baseline.

```{r, echo = FALSE}
##EVENT INCIDENCE BY HOUSEHOLD INCOME
xtabs(~income + event, data = surv.df)
```


Incident diabetes by baseline BMI group status.

```{r, echo =  FALSE}
xtabs(~bmi_group + event, data = surv.df)
```



# Stratified Cox Models

First we set the reference levels for factors. The reference level for insurance status is the uninsured level. For race we use the non-hispanic whites as the reference level, for BMI group we use BMI less than 30, and for household income we use the the $50,000 - 99,0000 group as it is the median and mode for household income.

```{r, echo = FALSE}
surv.df$insured = relevel(surv.df$insured, ref = "No")
levels(surv.df$insured)

surv.df$race = relevel(surv.df$race, ref = "Non-Hispanic White/Caucausian")
levels(surv.df$race)

surv.df$bmi_group = relevel(surv.df$bmi_group, ref = "<30")
levels(surv.df$bmi_group)

surv.df$income = relevel(surv.df$income, ref = "50-99")
levels(surv.df$income)
```

## Chi-Squared Test of Indpendence

We conduct a Pearson's Chi-Squared Test of Independence for each pair of factors. All pairs of factors are dependent, which makes sense as all factors are highly related to socioeconomic status (see p-values below). 

_P-value for Chi-Squared Test of Independence for Race and Insurance Status_
Race and insurance status are not independent, and the Cramer's V indicates moderately strong association between the two covariates.

```{r}
table(surv.df$race, surv.df$insured)
chisq.test(table(surv.df$race, surv.df$insured))
```
```{r}
cramersV(table(surv.df$race, surv.df$insured))
```

_P-value for Chi-Squared Test of Independence for Race and Baseline Household Income_

Race and income  are not independent, and the Cramer's V indicates a strong association between the two covariates.
```{r}
table(surv.df$race, surv.df$income)
chisq.test(table(surv.df$race, surv.df$income))
```
```{r}
cramersV(table(surv.df$race, surv.df$income))
```

_P-value for Chi-Squared Test of Independence for Race and BMI Group_

Race and BMI are not independent, with Cramer's V indicating a moderate association between the two covariates. 

```{r}
table(surv.df$race, surv.df$bmi_group)
chisq.test(table(surv.df$race, surv.df$bmi_group))
```
```{r}
cramersV(table(surv.df$race, surv.df$bmi_group))
```

_P-value for Chi-Squared Test of Independence for Insurance Status and Baseline Household Income_

Insurance status and household income are not independent, with Cramer's V indicating a moderately strong association between the two covariates. 
```{r}
table(surv.df$insured, surv.df$income)
chisq.test(table(surv.df$insured, surv.df$income))
```

```{r}
cramersV(table(surv.df$insured, surv.df$income))
```

_P-value for Chi-Squared Test of Independence for Insurance Status and BMI Group_

Insurance status and household income are not independent, with Cramer's V indicating a weak association between the two covariates. 

```{r}
table(surv.df$insured, surv.df$bmi_group)
chisq.test(table(surv.df$insured, surv.df$bmi_group))
```

```{r}
cramersV(table(surv.df$insured, surv.df$bmi_group))
```

_P-value for Chi-Squared Test of Independence for Baseline Household Income and BMI Group_

BMI and household income are not independent, with Cramer's V indicating a moderately weak association between the two covariates. 

```{r}
table(surv.df$income, surv.df$bmi_group)
chisq.test(table(surv.df$income, surv.df$bmi_group))
```
```{r}
cramersV(table(surv.df$income, surv.df$bmi_group))
```


## Stratified Cox Model with Only Health Insurance Status

We build a stratified Cox model with diabetes survival depending only on health insurance status and stratified by age group. To assess the validity of this output, we proceed to model diagnostics to assess the proportional hazards and no-interaction assumptions.

```{r insurance only, echo = FALSE}
##STRATIFIED COX MODEL WITH INSURANCE STRATIFIED BY AGE GROUP

#BUILD MODEL
dbts.mod.ins= coxph(surv_obj ~ insured + strata(age_group), 
                    data = surv.df)
summary(dbts.mod.ins)

```

We assess the proportional hazards assumption for the above model using a Goodness-of-Fit test based on the Schoenfeld residuals (Grambsch & Therneau, 1994). The proportional hazards assumption is met since the p-value is not significant ($p = 0.20$). Using the plot of the residuals below, we can see that the hazard ratio appears constant over time.

```{r SchoenResid_insured, echo = FALSE}
##PH ASSUMPTION
cox.zph(dbts.mod.ins)
plot(cox.zph(dbts.mod.ins), 
     main = "Scaled Schoenfeld Residuals (~insured + strata(age_group)")
```

We assess the no-interaction assumption for the age strata and insurance status. The coefficient estimates of separate models by age strata do not appear greatly different. We confirm that there is no significant interaction between the strata and insurance status with a Chi-Squared test of model difference with 2 degrees of freedom. 

```{r, echo = FALSE}
##NO INTERACTION ASSUMPTION
#The coefficients do not appear very different
coef(coxph(surv_obj ~ insured, 
      data = surv.df[surv.df$age_group == "42_45",]))
coef(coxph(surv_obj ~ insured, 
      data = surv.df[surv.df$age_group == "46_49",]))
coef(coxph(surv_obj ~ insured, 
      data = surv.df[surv.df$age_group == "50_53",]))

#stratified interaction model w/o main effect of age group
dbts.mod.interaction = coxph(surv_obj ~ insured*age_group - age_group + strata(age_group),
      data = surv.df)
dbts.mod.interaction

#interaction not significant
anova(dbts.mod.interaction, dbts.mod.ins)

```
With all model assumptions met, we interpret the model. The hazard ratio for those with insurance compared to those without insurance is 0.375 which appears significant at $\alpha = 0.001$ ($p < 0.0001$), suggesting that risk for diabetes is 62.5% lower for insured than for uninsured American women.

## Stratified Cox Model with Insurance Status, Race, BMI, and Household Income

We construct a stratified Cox model of diabetes survival depending on insurance status, race, BMI, and household income stratified by age group. In this model, we find that, after adjusting for race, baseline household income, baseline BMI group, and stratifying by age group, the effect of insurances status on diabetes survival is not significant ($p = 0.182$). All other factors appear significant, with all minority groups experiencing higher risk for diabetes incidence than non-Hispanic whites after adjusting for insurance, income, BMI, and age. Yearly household incomes less than $50,000 are at higher risk for diabetes incidence than those at higher income levels, and higher BMI groups are also at higher risk for diabetes incidence. We will assess the validity of this model, before further interpretation of the hazard ratios.

```{r all factors, echo = FALSE}
##STRATIFIED COX MODEL WITH INSURANCE, RACE, INCOME, and BMI group STRATIFIED BY AGE
#BUILD MODEL
dbts.mod.all = coxph(surv_obj ~ insured + race + income + bmi_group + strata(age_group), data = surv.df)
summary(dbts.mod.all)
```

We assess the proportional hazards assumption for the above model using a Goodness-of-Fit test based on the Schoenfeld residuals (Grambsch & Therneau, 1994). All p-values are greater than $\alpha = 0.05$, so the proportional hazards assumption is met for the model.

```{r SchoenResid_insured_race_income_bmi, echo = FALSE}
#PH ASSUMPTION
cox.zph(dbts.mod.all)
par(mfrow = c(2,2))
plot(cox.zph(dbts.mod.all))
```

Fitting separate proportional hazards models for each strata, it appears that the estimated coefficients are similar. We fit a larger model that includes the interaction between the strata and the covariates, and, using a Chi-squared test for model differences between the larger model and the previously estimated model, we evaluate the no-interaction assumption for the strata and the covariates. We find no significant interaction between the strata and the covariates. The no-interaction assumption is met.

```{r, echo = FALSE}
#NO INTERACTION ASSUMPTION
coef(coxph(surv_obj ~ insured + race + income + bmi_group, 
      data =surv.df[surv.df$age_group == "42_45",]))
coef(coxph(surv_obj ~ insured + race + income + bmi_group, 
      data = surv.df[surv.df$age_group == "46_49",]))
coef(coxph(surv_obj ~ insured + race + income + bmi_group, 
      data = surv.df[surv.df$age_group == "50_53",]))

dbts.mod.all.interaction = coxph(surv_obj ~ (insured + race + income + bmi_group)*age_group - age_group + strata(age_group), data = surv.df)

#chi-squared test for significance of interaction
anova(dbts.mod.all.interaction, dbts.mod.all)
```

## Stratified Cox Model with Race, BMI, and Household Income

We finally construct a model with only the significant factors found in the previous model, household income and race, and stratified by age. Using a Chi-Squared test of model difference, we find that the model including insurance status is not significantly better than the model without insurance status. A Chi-Squared test of deviances with 14 degrees of freedom finds no significant difference between the model including an interaction between age group and the covariates, so the no interaction assumption for stratified Cox models is also satisfied.

```{r reduced model, echo = FALSE}
##STRATIFIED COX MODEL WITH RACE, BMI, AND INCOME, AND STRATIFIED BY AGE
#BUILD MODEL
dbts.mod.noinsure = coxph(surv_obj ~ race + income + bmi_group + strata(age_group), data = surv.df)
dbts.mod.noinsure

#LRT Chi-Sq test for models
anova(dbts.mod.noinsure, dbts.mod.all)
```

We assess the proportional hazards assumption for the above model using a Goodness-of-Fit test based on the Schoenfeld residuals (Grambsch & Therneau, 1994). The proportional hazards assumption is met for all factors since there are no significant p-values at $\alpha = 0.05$. 

```{r ShoenResid_race_income_bmi, echo = FALSE}
#PH ASSUMPTION
cox.zph(dbts.mod.noinsure)
par(mfrow = c(2,2))
plot(cox.zph(dbts.mod.noinsure))
```

We assess the no-interaction assumption between the strata and the covariates using a Chi-squared test of model difference with 18 degrees of freedom. We find no significant interaction between the strata and the covariates, race, income, and BMI group.

```{r, echo = FALSE}
#NO INTERACTION ASSUMPTION
coef(coxph(surv_obj ~ race + income + bmi_group, data = surv.df[surv.df$age_group == "42_45",]))
coef(coxph(surv_obj ~ race + income + bmi_group, data = surv.df[surv.df$age_group == "46_49",]))
coef(coxph(surv_obj ~ race + income + bmi_group, data = surv.df[surv.df$age_group == "50_53",]))

dbts.mod.noinsure.interaction = coxph(surv_obj ~ (race + income + bmi_group)*age_group - age_group + strata(age_group), data = surv.df)

#Chi-Squared test of deviance
anova(dbts.mod.noinsure.interaction, dbts.mod.noinsure)

```

We find that the hazard ratios for diabetes incidence are highest for Hispanic women, with 195% higher risk for diabetes incidence after adjusting for household income, BMI group, and stratifying by age ($p < 0.0001$). After adjusting for household income, BMI group, and stratifying by age, all racial groups are at significantly higher risk than their non-Hispanic white counterparts. The risk for diabetes is significantly higher for groups with yearly household incomes less than $50,000 after adjusting for race, BMI, and stratifying by age. Notably, those with a BMI higher than 40 are at 592.92% higher risk and those with a BMI between 30 and 40 are at 265.9% higher risk for diabetes incidence than those with a BMI less than 30 after adjusting for race, household income, and stratified by race.

```{r selected model summary, echo = FALSE}
#MODEL SUMMARY
dbts.mod.noinsure
```

<!-- ## Stratified Cox Model with Covariate Interactions -->

<!-- Race and household income may be confounding in their effects on health status, since certain race/ethnic groups are over-represented in lower income groups. We assess whether there is a significant interaction between these two covariates after adjusting for BMI and stratifying by age group. The model shows no significant interactions between any race or income group. Low significance of interaction groups may be due to very few replications of factor combinations, which results in a convergence warning in the model fitting.  -->

<!-- ```{r, echo = FALSE} -->
<!-- coxph(surv_obj ~ insured + race*income + bmi_group + strata(age_group), data = surv.df) -->
<!-- ``` -->

# Stratified Cox Model with Insurance Status, Race, and BMI

To investigate which factor adjustment makes the effect of insurance status insignificant, we build separate stratified Cox models with race and one of the other factors: race, household income, or BMI group. From the models below, we see that adjusting for baseline household income makes the effect of insurance status insignificant. 

__CPH Model for Insurance and Race__

```{r, echo = FALSE}
coxph(surv_obj ~ insured + race + strata(age_group), data = surv.df)

cox.zph(coxph(surv_obj ~ insured + race + strata(age_group), data = surv.df))
```

__CPH Model for Insurance and Baseline Household Income__
```{r, echo = FALSE}
coxph(surv_obj ~ insured + income + strata(age_group), data = surv.df)
cox.zph(coxph(surv_obj ~ insured + income + strata(age_group), data = surv.df))
```

```{r, echo = FALSE}
coxph(surv_obj ~ insured + strata(income) + strata(age_group), data = surv.df)
cox.zph(coxph(surv_obj ~ insured + strata(income) + strata(age_group), data = surv.df))

```

__CPH Model for Insurance and BMI Group__
```{r, echo = FALSE}
coxph(surv_obj ~ insured + bmi_group + strata(age_group), data = surv.df)
cox.zph(coxph(surv_obj ~ insured + bmi_group + strata(age_group), data = surv.df))
```

We then consider a model with insurance status, race, and BMI group, stratified by age. We find that after adjusting for race, BMI group, and age, those with insurance verse those without insurance are at significantly lower risk for diabetes incidence (HR = 0.581, $p = 0.126$). We proceed to our model assumptions to validate this result. 

_CPH Model for Insurance, Race, and BMI Group__
```{r, echo = FALSE}
dbts.mod.noincome = coxph(surv_obj ~ insured + bmi_group + race + strata(age_group), data = surv.df)

summary(dbts.mod.noincome)
```

First, we assess the proportional hazards assumption for the above model using a Goodness-of-Fit test based on the Schoenfeld residuals (Grambsch & Therneau, 1994). The proportional hazards assumption is met for all factors since there are no significant p-values at $\alpha = 0.05$.

```{r ShoenResid_insured_bmi_race, echo = FALSE}
#PH ASSUMPTION
cox.zph(dbts.mod.noincome)
par(mfrow = c(2,2))
plot(cox.zph(dbts.mod.noincome))
```

We assess the no-interaction assumption between the strata and the covariates using a Chi-squared test of model difference with 14 degrees of freedom. We find no significant interaction between the strata and the covariates, insurance, race, and BMI group.

```{r, echo = FALSE}
#NO INTERACTION ASSUMPTION
dbts.mod.noincome.interaction = coxph(surv_obj ~ (insured + race + bmi_group)*age_group - age_group + strata(age_group), data = surv.df)

#Chi-Squared test of deviance
anova(dbts.mod.noincome.interaction, dbts.mod.noincome)
```

# Kaplan-Meier Curves

## Survival Curves by Insurance Status For Each Age Group

We construct separate Kaplan-Meier curves for the effect of health insurance by each of the age strata and use a log-rank test to detect significant differences between the observed survival curves. For those who entered the study between ages 42 to 45 and between ages 69 to 49 the risk for diabetes for those without insurance is significantly lower than for those with insurance ($p = 0.0005$ for ages 42 to 45, $p = 0.0001$ for ages 46-49). For women who entered the study between ages 50 and 53, there is no significant difference between the survival curves for those with insurance versus those without ($p = 0.3$).


```{r SurvCurve_42_45_insurance, echo = FALSE}
km42_45 = survfit(surv_obj ~ insured, 
                  data = surv.df[surv.df$age_group=="42_45",])
ggsurvplot(km42_45, 
           data = surv.df[surv.df$age_group=="42_45",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (42-45 Age Group)",
           ylim = c(0.6, 1.0),
           legend.title = "Insurance Status",
           legend.labs = c("Uninsured", "Insured"),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           ggtheme = theme_classic2())

survdiff(surv_obj ~ insured, 
         data = surv.df[surv.df$age_group=="42_45",])


```

```{r SurvCurve_46_49_insurance, echo = FALSE}
km46_49 = survfit(surv_obj ~ insured, 
                  data = surv.df[surv.df$age_group=="46_49",])
ggsurvplot(km46_49, 
           data = surv.df[surv.df$age_group=="46_49",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (46-49 Age Group)",
           ylim = c(0.6, 1.0),
           legend.title = "Insurance Status",
           legend.labs = c("Uninsured", "Insured"),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           ggtheme = theme_classic2(base_family = "Times"))

survdiff(surv_obj ~ insured, 
                  data = surv.df[surv.df$age_group=="46_49",])
```

```{r SurvCurve_50_53_insurance, echo = FALSE}
km50_53 = survfit(surv_obj ~ insured, 
                  data = surv.df[surv.df$age_group=="50_53",])
ggsurvplot(km50_53, 
           data = surv.df[surv.df$age_group=="50_53",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (50-53 Age Group)",
           ylim = c(0.6, 1.0),
           legend.title = "Insurance Status",
           legend.labs = c("Uninsured", "Insured"),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           ggtheme = theme_classic2())

survdiff(surv_obj ~ insured, 
                  data = surv.df[surv.df$age_group=="50_53",])
```


## Survival Curves by Race For Each Age Group

We construct separate Kaplan-Meier curves for the effect of race by each of the age strata and use a log-rank test to detect significant differences between the observed survival curves. For all age groups, survival curves significantly differ by race ($p < 0.0001$ for ages 43-46, $p < 0.0001$ for ages 46-49, $p < 0.0001$ for ages 50-53). Differences appear especially large for Hispanic women.

```{r SurvCurve_42_45_race, echo = FALSE}
km42_45 = survfit(surv_obj ~ race, 
                  data = surv.df[surv.df$age_group=="42_45",])
ggsurvplot(km42_45, 
           data = surv.df[surv.df$age_group=="42_45",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (42-45 Age Group)",
           ylim = c(0.6, 1.0),
           legend.title = "Race",
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.labs = c("White", "Black/African American", "Chinese/Chinese American", "Japanese/Japanese American", "Hispanic"),
           ggtheme = theme_classic2())

```

```{r}
survdiff(surv_obj ~ race, 
         data = surv.df[surv.df$age_group=="42_45",])
```

```{r}
pairwise_survdiff(surv_obj ~ race, 
         data = surv.df[surv.df$age_group=="42_45",])
```


```{r SurvCurve_46_49_race, echo = FALSE}
km46_49 = survfit(surv_obj ~ race, 
                  data = surv.df[surv.df$age_group=="46_49",])
ggsurvplot(km46_49, 
           data = surv.df[surv.df$age_group=="46_49",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (46-49 Age Group)",
           ylim = c(0.6, 1.0),
           legend.title = "Race",
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.labs = c("White", "Black/African American", "Chinese/Chinese American", "Japanese/Japanese American", "Hispanic"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ race, 
         data = surv.df[surv.df$age_group=="46_49",])
```

```{r}
pairwise_survdiff(surv_obj ~ race, 
         data = surv.df[surv.df$age_group=="46_49",])
```



```{r SurvCurve_50_53_race, echo = FALSE}
km50_53 = survfit(surv_obj ~ race, 
                  data = surv.df[surv.df$age_group=="50_53",])
ggsurvplot(km50_53, 
           data = surv.df[surv.df$age_group=="50_53",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (50-53 Age Group)",
           ylim = c(0.6, 1.0),
           legend.title = "Race",
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.labs = c("White", "Black/African American", "Chinese/Chinese American", "Japanese/Japanese American", "Hispanic"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ race, 
         data = surv.df[surv.df$age_group=="50_53",])
```

```{r}
pairwise_survdiff(surv_obj ~ race, 
         data = surv.df[surv.df$age_group=="50_53",])
```

## Survival Curves by Baseline Household Income for Each Age Group

We construct separate Kaplan-Meier curves for the effect of baseline household income by each of the age strata and use a log-rank test to detect significant differences between the observed survival curves. For all age groups, survival curves significantly differ by race ($p < 0.0001$ for ages 43-46, $p < 0.0001$ for ages 46-49, $p = 0.0002$ for ages 50-53). Among all age groups, these differences appear especially large in the lowest income group which experiences a faster decline in survival probability than other groups. This effect is particularly pronounced in women who entered the study between ages 50 and 53.

```{r SurvCurve_42_45_income, echo = FALSE}

##relevel for plotting
surv.df$income = factor(surv.df$income, levels = c("<20", "20-49", "50-99", "100+"))

km42_45 = survfit(surv_obj ~ income, 
                  data = surv.df[surv.df$age_group=="42_45",])
ggsurvplot(km42_45, 
           data = surv.df[surv.df$age_group=="42_45",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (42-45 Age Group)",
           ylim = c(0.6, 1.0),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.title = "Income in thousands of dollars",
           legend.labs = c("<20", "20-49", "50-99", "100+"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ income, 
         data = surv.df[surv.df$age_group=="42_45",])
```

```{r}
pairwise_survdiff(surv_obj ~ income, 
         data = surv.df[surv.df$age_group=="42_45",])
```


```{r SurvCurve_46_49_income, echo = FALSE}
km46_49 = survfit(surv_obj ~ income, 
                  data = surv.df[surv.df$age_group=="46_49",])
ggsurvplot(km46_49, 
           data = surv.df[surv.df$age_group=="46_49",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (46-49 Age Group)",
           ylim = c(0.6, 1.0),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.title = "Income in thousands of dollars",
           legend.labs = c("<20", "20-49", "50-99", "100+"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ income, 
         data = surv.df[surv.df$age_group=="46_49",])
```

```{r}
pairwise_survdiff(surv_obj ~ income, 
         data = surv.df[surv.df$age_group=="46_49",])
```


```{r SurvCurve_50_53_income, echo = FALSE}
km50_53 = survfit(surv_obj ~ income, 
                  data = surv.df[surv.df$age_group=="50_53",])
ggsurvplot(km50_53, 
           data = surv.df[surv.df$age_group=="50_53",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (50-53 Age Group)",
           ylim = c(0.6, 1.0),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.title = "Income in thousands of dollars",
           legend.labs = c("<20", "20-49", "50-99", "100+"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ income, 
         data = surv.df[surv.df$age_group=="50_53",])
```

```{r}
pairwise_survdiff(surv_obj ~ income, 
         data = surv.df[surv.df$age_group=="50_53",],
         p.adjust.method = "bonferroni")
```

## Survival Curves by BMI Group for Each Age Group

We construct separate Kaplan-Meier curves for the effect of baseline BMI for each of the age strata and use a log-rank test to detect significant differences between the observed survival curves. For all age groups, survival curves significantly differ by race ($p < 0.0001$ for ages 43-46, $p < 0.0001$ for ages 46-49, $p < 0.0001$ for ages 50-53). Across all age groups, higher BMI groups appear to experience a faster decline in survival probability.

```{r SurvCurve_42_45_bmi, echo = FALSE}

km42_45 = survfit(surv_obj ~ bmi_group, 
                  data = surv.df[surv.df$age_group=="42_45",])
ggsurvplot(km42_45, 
           data = surv.df[surv.df$age_group=="42_45",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (42-45 Age Group)",
           ylim = c(0.6, 1.0),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.title = "BMI Group",
           legend.labs = c("<30", "30_40", "40+"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ bmi_group, 
         data = surv.df[surv.df$age_group=="42_45",])
```

```{r}
pairwise_survdiff(surv_obj ~ bmi_group, 
         data = surv.df[surv.df$age_group=="42_45",])
```


```{r SurvCurve_46_49_bmi, echo = FALSE}
km46_49 = survfit(surv_obj ~ bmi_group, 
                  data = surv.df[surv.df$age_group=="46_49",])
ggsurvplot(km46_49, 
           data = surv.df[surv.df$age_group=="46_49",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (46-49 Age Group)",
           ylim = c(0.6, 1.0),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.title = "BMI Group",
           legend.labs = c("<30", "30_40", "40+"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ bmi_group, 
         data = surv.df[surv.df$age_group=="46_49",])
```

```{r}
pairwise_survdiff(surv_obj ~ bmi_group, 
         data = surv.df[surv.df$age_group=="46_49",])
```


```{r SurvCurve_50_53_bmi, echo = FALSE}
km50_53 = survfit(surv_obj ~ bmi_group, 
                  data = surv.df[surv.df$age_group=="50_53",])
ggsurvplot(km50_53, 
           data = surv.df[surv.df$age_group=="50_53",], 
           conf.int=FALSE, 
           xlab = "Time (t)", 
           ylab = "Survival Probability", 
           title = "Survival Function for Time to Diabetes (50-53 Age Group)",
           ylim = c(0.6, 1.0),
           risk.table = TRUE,
           risk.table.height = .5,
           break.x.by = 1,
           fontsize = 3,
           legend = "top",
           legend.title = "BMI Group",
           legend.labs = c("<30", "30_40", "40+"),
           ggtheme = theme_classic2())
```

```{r}
survdiff(surv_obj ~ bmi_group, 
         data = surv.df[surv.df$age_group=="50_53",])
```

```{r}
pairwise_survdiff(surv_obj ~ bmi_group, 
         data = surv.df[surv.df$age_group=="50_53",])
```


## Notes:
- Forward-backward selection yields the previously discovered model (the model without insurance) https://mathweb.ucsd.edu/~rxu/math284/model_sel.pdf
- If we don't adjust for income, the effect of insurance remains significant, even after adjusting for race, BMI, and age.
- If we don't stratify by age the income variable fails to meet the proportional hazards assumption. Why would this be?



